Verification of a Deterministic Random Bits Generator

### 2015-12-24 paper/paper.tex
paper: various style changes

\subsection{Machine-Checked Proofs}
I did the verification in Coq, an open-sourced proof assistant. The C code is compiled into its abstract syntax tree (AST) in Coq using the CompCert certified (machine-checked) compiler. I wrote functional specifications (functional programs) in Coq that closely match \stdtitle{NIST SP 800-90A}. Using the functional programs, I created API specifications using the VST framework. API specifications reason about how the AST deals with memory and local variables. I then prove the AST correct with respect to the API specifications.

Of what the tasks I mentioned, the ones that must be trusted are the translation of \stdtitle{NIST SP 800-90A} into functional specifications and the creation of API specifications from those functional programs. Everything else is machine-checked. Can the functional and API specifications be trusted? For the latter, I hope to make a convincing claim by the end of this paper that as long as the functional specifications are correct, the corresponding API specifications will be, too. As for the functional specifications, first of all, they are small, easy-to-reason-about functional programs. I also ran the programs on some test inputs from NIST and verified that the output matches. While this is already a huge step forward from a C program, as a functional program in Coq is less error-prone, it is still unsatisfying that at the functional specification level, I had to resort to unit tests. Thankfully, there is a parallel project by Katherine Ye ‘16 that will prove cryptographic properties of the functional specifications. The fact that this project can be done in parallel to (and mostly independent from) the verification of the C implementation shows the power and flexibility of VST.

@@

There have been some other verification efforts to verify a DRBG, albeit without using a functional specification. As an example, \cite{prng} verified Android’s pseudorandom number generator. This verification is done on a Java implementation, and is a verification on information flow, not on a functional specification. That is, the project proves that all the bytes of entropy are used in constructing the secret internal state. Proving an implementation against a functional specification is arguably better, as it allows the functional specification to serve as a mathematical representation of the more complex C (or Java) implementation. This functional specification can then be shown to have desired cryptographic properties, including the fact that all bytes of entropy are being properly consumed.

@@

Previous efforts in the VST project to verify SHA-256 \cite{sha} and HMAC \cite{hmac} have done so using \stdtitle{OpenSSL 0.9.1c}. A natural choice for the DRBG verification would be to also use the same \stdtitle{OpenSSL} library. Unfortunately, the DRBG implementation in \stdtitle{OpenSSL} makes heavy use of function pointers to encapsulate the differences between each type of DRBG. While Verifiable C is capable of reasoning about function pointers, the proof automation system is not yet robust enough when dealing with function pointers and so such proofs will be more complex. \stdtitle{mbedTLS}, on the other hand, does not make use of function pointers except to deal with entropy. Therefore, I chose to use the implementation of DRBG found in \stdtitle{mbedTLS 2.1.1}.

@@

\section{DRBG Algorithm}
The DRBG algorithm contains four public functions: \lstinline{instantiate}, \lstinline{reseed}, \lstinline{generate}, and \lstinline{uninstantiate}. I will focus on the first three, as \lstinline{uninstantiate} simply involves freeing memory.

\subsection{The DRBG State}

@@

Note that \lstinline{Z} is Coq’s representation of arbitrary-precision integers, and is used here to represent a byte. That is, the value of the integer must be between $0$ and $255$, inclusive. I could have represented each byte as a tuple of a \lstinline{Z} and a proof that the value of that \lstinline{Z} is in the byte range, keeping that fact separate allowed me to write a functional specification that closely resembles \stdtitle{NIST SP 800-90A}, and the fact that if the input state contains bytes, then the output will also contain bytes can easily be provided as separate lemmas about the individual functions.

\subsection{The Update Function}
Before showing any of the public function, I must first define a helper function, the \lstinline{update} function. See \stdtitle{NIST SP 800-90A \S 10.1.2.2} for the corresponding NIST specification. The \lstinline{update} function takes a \lstinline{key}, a \lstinline{value}, and an optional nonce and returns a new \lstinline{key} and a new \lstinline{value}. This function will be called at least once in each of the public functions, and the goal is to provide backtracking resistance. What this means is that the original \lstinline{key} cannot be computed from the new \lstinline{key} (note that the \lstinline{value}(s) is public anyways; see \ref{funcgenerate}).

@@

\begin{lstlisting}
    Definition DRBG_instantiate_function
      (instantiate_algorithm: list Z -> list Z -> list Z -> Z -> DRBG_working_state)
      (min_entropy_length max_entropy_length: Z)
      (get_nonce: unit -> list Z)
      (highest_supported_security_strength: Z)
      (max_personalization_string_length: Z)
      (prediction_resistance_supported: bool)
      (entropy_stream: ENTROPY.stream)
      (requested_instantiation_security_strength: Z)
      (prediction_resistance_flag: bool)
      (personalization_string: list Z)
      : ENTROPY.result DRBG_state_handle :=
      if requested_instantiation_security_strength >? highest_supported_security_strength
        then ENTROPY.error ENTROPY.generic_error entropy_stream
      else match prediction_resistance_flag, prediction_resistance_supported with
             | true, false => ENTROPY.error ENTROPY.generic_error entropy_stream
             | _,_ =>
               if (Zlength personalization_string) >? max_personalization_string_length
                 then ENTROPY.error ENTROPY.generic_error entropy_stream
               else
                 let security_strength :=
                   if requested_instantiation_security_strength <=? 14 then Some 14
                   else if requested_instantiation_security_strength <=? 16 then Some 16
                   else if requested_instantiation_security_strength <=? 24 then Some 24
                   else if requested_instantiation_security_strength <=? 32 then Some 32
                   else None in
                 match security_strength with
                   | None => ENTROPY.error ENTROPY.generic_error entropy_stream
                   | Some security_strength =>
                   match (get_entropy security_strength min_entropy_length max_entropy_length
                         prediction_resistance_flag entropy_stream) with
                     | ENTROPY.error e s' => ENTROPY.error ENTROPY.catastrophic_error s'
                     | ENTROPY.success entropy_input entropy_stream =>
                       match (get_entropy (security_strength/2) (min_entropy_length/2)
                             (max_entropy_length/2) prediction_resistance_flag entropy_stream) with
                         | ENTROPY.error e s' => ENTROPY.error ENTROPY.catastrophic_error s'
                         | ENTROPY.success nonce entropy_stream =>
                           let initial_working_state := instantiate_algorithm entropy_input nonce
                             personalization_string security_strength in
                           ENTROPY.success (initial_working_state, security_strength,
                             prediction_resistance_flag) entropy_stream
                       end
                   end
                 end
           end.
\end{lstlisting}

Note that this is just the “outer” function that does error checking on the parameters, as mentioned earlier. Again, this is purely generic; different types of DRBG can share this piece of code. The actual algorithm for computing an initial working state is left as a dependency in the first parameter. All parameters before \lstinline{entropy_stream} are implementation specific constants. For example, \lstinline{prediction_resistance_supported} is a flag denoting whether or not this particular implementation can provide prediction resistance.

@@

\begin{lstlisting}
    Definition HMAC_DRBG_instantiate_algorithm (HMAC: list Z -> list Z -> list Z)
      (entropy_input nonce personalization_string: list Z) (security_strength: Z)
      : DRBG_working_state :=
      let seed_material := entropy_input ++ nonce ++ personalization_string in
      let key := initial_key in
      let value := initial_value in
      let (key, value) := HMAC_DRBG_update HMAC seed_material key value in
      let reseed_counter := 1 in
      (value, key, reseed_counter).
\end{lstlisting}

@@

\subsection{The Reseed Function}
There are times when the DRBG state must be reseeded with new bits of entropy. This can happen because too many bytes have been generated from the current bits of entropy, or because prediction resistance is requested. Reseeding with fresh entropy automatically provides prediction resistance, as the state at time $t + 1$ cannot be computed from the state at time $t$ alone - entropy is also needed. Note that reseeding is preferred over simply instantiating a new DRBG state because if the source of entropy fails silently (supposed the bits of entropy returned are not actually random), at least the DRBG still has some randomness left over from the old entropy.

@@

\begin{lstlisting}
    Definition DRBG_reseed_function
      (reseed_algorithm: DRBG_working_state -> list Z -> list Z -> DRBG_working_state)
      (min_entropy_length max_entropy_length: Z)
      (max_additional_input_length: Z)
      (entropy_stream: ENTROPY.stream)
      (state_handle: DRBG_state_handle)
      (prediction_resistance_request: bool)
      (additional_input: list Z)
      : ENTROPY.result DRBG_state_handle :=
      match state_handle with (working_state, security_strength, prediction_resistance_flag) =>
      if prediction_resistance_request && (negb prediction_resistance_flag)
        then ENTROPY.error ENTROPY.generic_error entropy_stream
      else
        if (Zlength additional_input) >? max_additional_input_length
          then ENTROPY.error ENTROPY.generic_error entropy_stream
        else
          match (get_entropy security_strength min_entropy_length max_entropy_length
                prediction_resistance_flag entropy_stream) with
            | ENTROPY.error _ s => ENTROPY.error ENTROPY.catastrophic_error s
            | ENTROPY.success entropy_input entropy_stream =>
              let new_working_state := reseed_algorithm working_state entropy_input additional_input in
              ENTROPY.success (new_working_state, security_strength,
                prediction_resistance_flag) entropy_stream
          end
      end.
\end{lstlisting}

@@

Note that both the \lstinline{reseed} and \lstinline{generate} functions allow users to pass in an additional stream of bytes. Users do not need to provide this (and implementations can choose to omit this feature), but it is meant to provide any additional bits of entropy or to provide personalization of this specific call.

@@

\begin{lstlisting}
    Definition HMAC_DRBG_reseed_algorithm (HMAC: list Z -> list Z -> list Z)
      (working_state: DRBG_working_state) (entropy_input additional_input: list Z)
      : DRBG_working_state :=
      match working_state with (v, key, _) =>
        let seed_material := entropy_input ++ additional_input in
        let (key, v) := HMAC_DRBG_update HMAC seed_material key v in
        let reseed_counter := 1 in
        (v, key, reseed_counter)
      end.
\end{lstlisting}

@@

\begin{lstlisting}
    Fixpoint DRBG_generate_function_helper
      (generate_algorithm: DRBG_working_state -> Z -> list Z -> DRBG_generate_algorithm_result)
      (reseed_function: ENTROPY.stream -> DRBG_state_handle -> bool -> list Z
        -> ENTROPY.result DRBG_state_handle)
      (entropy_stream: ENTROPY.stream)
      (state_handle: DRBG_state_handle)
      (requested_number_of_bytes: Z)
      (prediction_resistance_request: bool)
      (additional_input: list Z)
      (should_reseed: bool)
      (count: nat)
      : ENTROPY.result (list Z * DRBG_working_state) :=
      let result := if should_reseed then
                      match (reseed_function entropy_stream state_handle prediction_resistance_request
                        additional_input) with
                        | ENTROPY.success x entropy_stream => ENTROPY.success (x, []) entropy_stream
                        | ENTROPY.error e entropy_stream => ENTROPY.error e entropy_stream
                      end
                    else ENTROPY.success (state_handle, additional_input) entropy_stream in
      match result with
        | ENTROPY.error e s => ENTROPY.error e s
        | ENTROPY.success (state_handle, additional_input) entropy_stream =>
          match state_handle with (working_state, security_strength, prediction_resistance_flag) =>
            match generate_algorithm working_state requested_number_of_bytes additional_input with
              | generate_algorithm_reseed_required =>
                match count with
                  | O => ENTROPY.error ENTROPY.generic_error entropy_stream (* impossible *)
                  | S count' => (DRBG_generate_function_helper generate_algorithm reseed_function
                    entropy_stream state_handle requested_number_of_bytes prediction_resistance_request
                    additional_input true count')
                end
              | generate_algorithm_success x y => ENTROPY.success (x, y) entropy_stream
            end
          end
        end.

    Definition DRBG_generate_function
      (generate_algorithm: Z -> DRBG_working_state -> Z -> list Z -> DRBG_generate_algorithm_result)
      (reseed_function: ENTROPY.stream -> DRBG_state_handle -> bool -> list Z
        -> ENTROPY.result DRBG_state_handle)
      (reseed_interval: Z)
      (max_number_of_bytes_per_request: Z)
      (max_additional_input_length: Z)
      (entropy_stream: ENTROPY.stream)
      (state_handle: DRBG_state_handle)
      (requested_number_of_bytes requested_security_strength: Z)
      (prediction_resistance_request: bool)
      (additional_input: list Z)
      : ENTROPY.result (list Z * DRBG_state_handle) :=
      match state_handle with (working_state, security_strength, prediction_resistance_flag) =>
        if requested_number_of_bytes >? max_number_of_bytes_per_request
          then ENTROPY.error ENTROPY.generic_error entropy_stream
        else
          if requested_security_strength >? security_strength
            then ENTROPY.error ENTROPY.generic_error entropy_stream
          else
            if (Zlength additional_input) >? max_additional_input_length
              then ENTROPY.error ENTROPY.generic_error entropy_stream
            else
              if prediction_resistance_request && (negb prediction_resistance_flag)
                then ENTROPY.error ENTROPY.generic_error entropy_stream
              else
                match (DRBG_generate_function_helper (generate_algorithm reseed_interval)
                reseed_function entropy_stream state_handle requested_number_of_bytes
                prediction_resistance_request additional_input prediction_resistance_request 1%nat) with
                  | ENTROPY.error e s => ENTROPY.error e s
                  | ENTROPY.success (output, new_working_state) entropy_stream =>
                      ENTROPY.success (output, (new_working_state, security_strength,
                        prediction_resistance_flag)) entropy_stream
                end
      end.
\end{lstlisting}

Note that \lstinline{Fixpoint} is a recursive function where at least one parameter is structurally getting smaller. In this case, it is the last parameter, the natural number \lstinline{count}, which is initialized with \lstinline{1}, meaning that this recursive function is only allowed to go one level deep.

@@

    Definition HMAC_DRBG_generate_algorithm (HMAC: list Z -> list Z -> list Z) (reseed_interval: Z)
      (working_state: DRBG_working_state) (requested_number_of_bytes: Z) (additional_input: list Z)
      : DRBG_generate_algorithm_result :=
        match working_state with (v, key, reseed_counter) =>
          if reseed_counter >? reseed_interval then generate_algorithm_reseed_required
          else
            let (key, v) := match additional_input with
                              | [] => (key, v)
                              | _::_ => HMAC_DRBG_update HMAC additional_input key v
                            end in
            let (v, temp) := HMAC_DRBG_generate_helper_Z HMAC key v requested_number_of_bytes in
            let returned_bits := firstn (Z.to_nat requested_number_of_bytes) temp in
            let (key, v) := HMAC_DRBG_update HMAC additional_input key v in
            let reseed_counter := reseed_counter + 1 in
            generate_algorithm_success returned_bits (v, key, reseed_counter)
        end.
\end{lstlisting}

@@

Here, \lstinline{HMAC_Final} is the name of the \lstinline{final} function verified by previous efforts, and\\ \lstinline{mbedtls_md_hmac_finish} is the name mbedTLS expects. The API specification (how the function interacts with local variable and memory) for this function is almost a complete copy of the specification defined by the previous HMAC verification efforts.

@@

Recall that in the functional specification, I chose to represent the DRBG state as the following:

@@

When writing the API specification, I chose a representation that more closely resembles what is in \stdtitle{mbedTLS}. Naturally, \stdtitle{mbedTLS} represents the DRBG state as a struct. Note that the original \stdtitle{mbedTLS} code includes a function pointer for the entropy function as well as context to pass to the entropy function. As noted earlier, for the purposes of this project, I made the modifications to remove this function pointer.

@@

Lastly, I define the conversions between the abstract state and the functional state:

@@

One of the goals of my project is to serve as a bleeding-edge adopter of the VST framework and the HMAC API specification. As is the nature of bleeding-edge adoption, I encountered many difficulties, such as a proof tactic failing silently and not providing sufficient error messages. These difficulties are fixed as they get reported back to the developers of VST. In this section, I highlight one particular difficulty with the HMAC API specification.

\subsection{HMAC API Specification}

@@

This representation works in the sense that the verification of the \stdtitle{OpenSSL} functions went through. Unfortunately, for clients of this API specification, this representation is often not good enough. The biggest issue is that this representation is too low-level; it does not keep track of the key. Upon notified of this issue, the developers of the HMAC API specification were quickly able to provide a more abstract representation of the HMAC state, leading to the state machine representation shown earlier in \S \ref{hmac_interface}. The core of the verification code did not need to be changed, with some book-keeping added at the ends of translate between the new, more abstract representation to the more low-level implementation. While this issue was easily fixed, it would have been very hard to discover this issue without actually being clients of the API specification. This highlights the importance of having being the first client of the HMAC API specification.

@@

\begin{lstlisting}[language=C]
    unsigned char mbedtls_md_get_size( const mbedtls_md_info_t *md_info );

    int mbedtls_md_setup( mbedtls_md_context_t *ctx, const mbedtls_md_info_t *md_info, int hmac );

@@

    int mbedtls_md_hmac_reset( mbedtls_md_context_t *ctx );

@@

    void mbedtls_md_free( mbedtls_md_context_t *ctx );
\end{lstlisting}

@@

The \lstinline{md_info} field can mostly be ignored, as it is meant to be used to differentiate between the different types of message digest, and in this case, we are specializing the message digest to use HMAC and SHA-256. In order for clients of the message digest interface to have access to the \stdtitle{OpenSSL} HMAC context, I chose to have the \lstinline{hmac_ctx} pointer point to the \stdtitle{OpenSSL} HMAC context. This initialization happens in the \lstinline{mbedtls_md_setup} function:

@@

Lastly, the \lstinline{WITH} value \lstinline{kv} represents the pointer value to the SHA K vector, which is a global constant array in the SHA-256 implementation. For more information, see the \cite{sha}.

@@

Again, much like the \lstinline{update} function, the next three \lstinline{WITH} values are the pointer value to the DRBG state, the abstract DRBG state, and the concrete DRBG state. The only new property is in the \lstinline{PROP} part of the precondition, requiring the \lstinline{entropy_len} component to be 32. Note that in \stdtitle{NIST SP 800-90A} and consequently in the functional specification, this is a global, fixed constant, but \stdtitle{mbedTLS} varies this value based on the type of HMAC used and fixes the value upon instantiation. With HMAC256, this is fixed to be 32.

>>>>

\subsection{Machine-Checked Proofs}
I did the verification in Coq, an open-sourced proof assistant. The C code is compiled into its abstract syntax tree (AST) in Coq using the CompCert certified (machine-checked) compiler. I wrote functional specifications (functional programs) in Coq that closely match \stdtitle{NIST SP 800-90A}. Using the functional programs, I created API specifications using the VST framework. API specifications reason about how the AST deals with memory and local variables. I then proved the AST correct with respect to the API specifications.

Of what the tasks I mentioned, the ones that must be trusted are the translation of \stdtitle{NIST SP 800-90A} into functional specifications and the creation of API specifications from those functional programs. Everything else is machine-checked. Can the functional and API specifications be trusted? For the latter, I hope to make a convincing claim by the end of this paper that as long as the functional specifications are correct, the corresponding API specifications will be, too. As for the functional specifications, first of all, they are small, easy-to-reason-about, strongly typed functional programs. I also ran the programs on some test inputs from NIST and verified that the output matches. That said, it is still unsatisfying that at the functional specification level, I had to resort to unit tests. Thankfully, there is a parallel project by Katherine Ye ‘16 that will prove cryptographic properties of the functional specifications. The fact that this parallel project can be done in concurrent to (and mostly independent from) the verification of the C implementation shows the power and flexibility of VST.

@@

There have been some other verification efforts to verify a DRBG, albeit without using a functional specification. As an example, D{\"o}rre and Klebanov \cite{prng} verified Android’s pseudorandom number generator. This verification is done on a Java implementation, and is a verification on information flow, not on a functional specification. That is, the project proves that all the bytes of entropy are used in constructing the secret internal state, but not \textit{how} the bytes are used. Proving an implementation against a functional specification is arguably better, as it allows the functional specification to serve as a mathematical representation of the more complex C (or Java) implementation. This functional specification can then be shown to have desired cryptographic properties, including the fact that all bytes of entropy are being properly consumed.

@@

Previous efforts in the VST project to verify SHA-256 \cite{sha} and HMAC \cite{hmac} have done so using \stdtitle{OpenSSL 0.9.1c}. A natural choice for the DRBG verification would be to also use the same \stdtitle{OpenSSL} library. Unfortunately, the DRBG implementation in \stdtitle{OpenSSL} makes heavy use of function pointers to encapsulate the differences between each type of DRBG. While Verifiable C, the C program logic component of the VST framework, is capable of reasoning about function pointers, the proof automation system is not yet robust enough when dealing with function pointers and so such proofs will be more complex. \stdtitle{mbedTLS}, on the other hand, does not make use of function pointers except to deal with entropy. Therefore, I chose to use the implementation of DRBG found in \stdtitle{mbedTLS 2.1.1}.

@@

\section{DRBG Algorithm}
The DRBG algorithm contains four public functions: \lstinline{instantiate}, \lstinline{reseed}, \lstinline{generate}, and \lstinline{uninstantiate}. The HMAC-based DRBG also makes use of a helper function: \lstinline{update}. I will focus on all but \lstinline{uninstantiate}, as it simply involves freeing memory.

\subsection{The DRBG State} \label{funcstate}

@@

Note that \lstinline{Z} is Coq’s representation of arbitrary-precision integers, and is used here to represent a byte. That is, the value of the integer must be between $0$ and $255$, inclusive. While I could have represented each byte as a tuple of a \lstinline{Z} and a proof that the value of that \lstinline{Z} is in the byte range, keeping that fact separate allowed me to write a functional specification that closely resembles \stdtitle{NIST SP 800-90A}. The fact that if the input state contains bytes, then the output will also contain bytes can easily be provided as separate lemmas about the individual functions.

\subsection{The Update Function}
Before showing any of the public function, I must first define a helper function, the \lstinline{update} function. See \stdtitle{NIST SP 800-90A \S 10.1.2.2} for the corresponding NIST specification. The \lstinline{update} function takes a \lstinline{key}, a \lstinline{value}, and an optional nonce and returns a new \lstinline{key} and a new \lstinline{value}. This function will be called at least once in each of the public functions, and the goal is to provide backtracking resistance. What this means is that the original \lstinline{key} cannot be computed from the new \lstinline{key} (note that the \lstinline{value}(s) is public anyways; see \S \ref{funcgenerate}).

@@

\begin{lstlisting}
    Definition DRBG_instantiate_function
      (instantiate_algorithm: list Z -> list Z -> list Z -> Z -> DRBG_working_state)
      (min_entropy_length max_entropy_length: Z)
      (get_nonce: unit -> list Z)
      (highest_supported_security_strength: Z)
      (max_personalization_string_length: Z)
      (prediction_resistance_supported: bool)
      (entropy_stream: ENTROPY.stream)
      (requested_instantiation_security_strength: Z)
      (prediction_resistance_flag: bool)
      (personalization_string: list Z)
      : ENTROPY.result DRBG_state_handle :=
      if requested_instantiation_security_strength >? highest_supported_security_strength
        then ENTROPY.error ENTROPY.generic_error entropy_stream
      else match prediction_resistance_flag, prediction_resistance_supported with
             | true, false => ENTROPY.error ENTROPY.generic_error entropy_stream
             | _,_ =>
               if (Zlength personalization_string) >? max_personalization_string_length
                 then ENTROPY.error ENTROPY.generic_error entropy_stream
               else
                 let security_strength :=
                   if requested_instantiation_security_strength <=? 14 then Some 14
                   else if requested_instantiation_security_strength <=? 16 then Some 16
                   else if requested_instantiation_security_strength <=? 24 then Some 24
                   else if requested_instantiation_security_strength <=? 32 then Some 32
                   else None in
                 match security_strength with
                   | None => ENTROPY.error ENTROPY.generic_error entropy_stream
                   | Some security_strength =>
                   match get_entropy security_strength min_entropy_length max_entropy_length prediction_resistance_flag entropy_stream with
                     | ENTROPY.error e s' => ENTROPY.error ENTROPY.catastrophic_error s'
                     | ENTROPY.success entropy_input entropy_stream =>
                       match get_entropy (security_strength/2) (min_entropy_length/2) (max_entropy_length/2) prediction_resistance_flag entropy_stream with
                         | ENTROPY.error e s' => ENTROPY.error ENTROPY.catastrophic_error s'
                         | ENTROPY.success nonce entropy_stream =>
                           let initial_working_state := instantiate_algorithm entropy_input nonce
                             personalization_string security_strength in
                           ENTROPY.success (initial_working_state, security_strength,
                             prediction_resistance_flag) entropy_stream
                       end
                   end
                 end
           end.
\end{lstlisting}

Note that this is just the “outer” function that does error checking on the parameters, as mentioned earlier. This outer function is purely generic; different types of DRBG can share this piece of code. The actual algorithm for computing an initial working state is left as a dependency in the first parameter. All parameters before \lstinline{entropy_stream} are implementation specific constants. For example, \lstinline{prediction_resistance_supported} is a flag denoting whether or not this particular implementation can provide prediction resistance.

@@

\begin{lstlisting}
    Definition HMAC_DRBG_instantiate_algorithm (HMAC: list Z -> list Z -> list Z) (entropy_input nonce personalization_string: list Z) (security_strength: Z): DRBG_working_state :=
      let seed_material := entropy_input ++ nonce ++ personalization_string in
      let key := initial_key in
      let value := initial_value in
      let (key, value) := HMAC_DRBG_update HMAC seed_material key value in
      let reseed_counter := 1 in
      (value, key, reseed_counter).
\end{lstlisting}

@@

\subsection{The Reseed Function}
There are times when the DRBG state must be reseeded with new bits of entropy. This can happen because \lstinline{generate} has been called many times with the current bits of entropy, or because prediction resistance is requested. Reseeding with fresh entropy automatically provides prediction resistance, as the state at time $t + 1$ cannot be computed from the state at time $t$ alone - entropy is also needed. Note that reseeding is preferred over simply instantiating a new DRBG state because if the source of entropy fails silently (supposed the bits of entropy returned are not actually random), at least the DRBG still has some randomness left over from the old entropy.

@@

\begin{lstlisting}
    Definition DRBG_reseed_function
      (reseed_algorithm: DRBG_working_state -> list Z -> list Z -> DRBG_working_state)
      (min_entropy_length max_entropy_length: Z)
      (max_additional_input_length: Z)
      (entropy_stream: ENTROPY.stream)
      (state_handle: DRBG_state_handle)
      (prediction_resistance_request: bool)
      (additional_input: list Z)
      : ENTROPY.result DRBG_state_handle :=
      match state_handle with (working_state, security_strength, prediction_resistance_flag) =>
      if prediction_resistance_request && (negb prediction_resistance_flag)
        then ENTROPY.error ENTROPY.generic_error entropy_stream
      else
        if (Zlength additional_input) >? max_additional_input_length
          then ENTROPY.error ENTROPY.generic_error entropy_stream
        else
          match get_entropy security_strength min_entropy_length max_entropy_length prediction_resistance_flag entropy_stream with
            | ENTROPY.error _ s => ENTROPY.error ENTROPY.catastrophic_error s
            | ENTROPY.success entropy_input entropy_stream =>
              let new_working_state := reseed_algorithm working_state entropy_input additional_input in
              ENTROPY.success (new_working_state, security_strength,
                prediction_resistance_flag) entropy_stream
          end
      end.
\end{lstlisting}

@@

Note that both the \lstinline{reseed} and \lstinline{generate} functions allow users to pass in an additional sequence of bytes. Users do not need to provide this (and implementations can choose to omit this feature), but it is meant to provide any additional bits of entropy or to provide personalization of this specific call.

@@

\begin{lstlisting}
    Definition HMAC_DRBG_reseed_algorithm (HMAC: list Z -> list Z -> list Z) (working_state: DRBG_working_state) (entropy_input additional_input: list Z): DRBG_working_state :=
      match working_state with (v, key, _) =>
        let seed_material := entropy_input ++ additional_input in
        let (key, v) := HMAC_DRBG_update HMAC seed_material key v in
        let reseed_counter := 1 in
        (v, key, reseed_counter)
      end.
\end{lstlisting}

@@

\begin{lstlisting}
    Fixpoint DRBG_generate_function_helper
      (generate_algorithm: DRBG_working_state -> Z -> list Z -> DRBG_generate_algorithm_result)
      (reseed_function: ENTROPY.stream -> DRBG_state_handle -> bool -> list Z
        -> ENTROPY.result DRBG_state_handle)
      (entropy_stream: ENTROPY.stream)
      (state_handle: DRBG_state_handle)
      (requested_number_of_bytes: Z)
      (prediction_resistance_request: bool)
      (additional_input: list Z)
      (should_reseed: bool)
      (count: nat)
      : ENTROPY.result (list Z * DRBG_working_state) :=
      let result := if should_reseed then
                      match reseed_function entropy_stream state_handle prediction_resistance_request additional_input with
                        | ENTROPY.success x entropy_stream => ENTROPY.success (x, []) entropy_stream
                        | ENTROPY.error e entropy_stream => ENTROPY.error e entropy_stream
                      end
                    else ENTROPY.success (state_handle, additional_input) entropy_stream in
      match result with
        | ENTROPY.error e s => ENTROPY.error e s
        | ENTROPY.success (state_handle, additional_input) entropy_stream =>
          match state_handle with (working_state, security_strength, prediction_resistance_flag) =>
            match generate_algorithm working_state requested_number_of_bytes additional_input with
              | generate_algorithm_reseed_required =>
                match count with
                  | O => ENTROPY.error ENTROPY.generic_error entropy_stream (* impossible *)
                  | S count' => DRBG_generate_function_helper generate_algorithm reseed_function entropy_stream state_handle requested_number_of_bytes prediction_resistance_request additional_input true count'
                end
              | generate_algorithm_success x y => ENTROPY.success (x, y) entropy_stream
            end
          end
        end.

    Definition DRBG_generate_function
      (generate_algorithm: Z -> DRBG_working_state -> Z -> list Z -> DRBG_generate_algorithm_result)
      (reseed_function: ENTROPY.stream -> DRBG_state_handle -> bool -> list Z
        -> ENTROPY.result DRBG_state_handle)
      (reseed_interval: Z)
      (max_number_of_bytes_per_request: Z)
      (max_additional_input_length: Z)
      (entropy_stream: ENTROPY.stream)
      (state_handle: DRBG_state_handle)
      (requested_number_of_bytes requested_security_strength: Z)
      (prediction_resistance_request: bool)
      (additional_input: list Z)
      : ENTROPY.result (list Z * DRBG_state_handle) :=
      match state_handle with (working_state, security_strength, prediction_resistance_flag) =>
        if requested_number_of_bytes >? max_number_of_bytes_per_request
          then ENTROPY.error ENTROPY.generic_error entropy_stream
        else
          if requested_security_strength >? security_strength
            then ENTROPY.error ENTROPY.generic_error entropy_stream
          else
            if (Zlength additional_input) >? max_additional_input_length
              then ENTROPY.error ENTROPY.generic_error entropy_stream
            else
              if prediction_resistance_request && (negb prediction_resistance_flag)
                then ENTROPY.error ENTROPY.generic_error entropy_stream
              else
                match DRBG_generate_function_helper (generate_algorithm reseed_interval) reseed_function entropy_stream state_handle requested_number_of_bytes prediction_resistance_request additional_input prediction_resistance_request 1%nat with
                  | ENTROPY.error e s => ENTROPY.error e s
                  | ENTROPY.success (output, new_working_state) entropy_stream =>
                      ENTROPY.success (output, (new_working_state, security_strength, prediction_resistance_flag)) entropy_stream
                end
      end.
\end{lstlisting}

Note that \lstinline{Fixpoint} is a recursive function where at least one parameter is structurally getting smaller, allowing Coq to figure out the proof of termination automatically. In this case, it is the last parameter, the natural number \lstinline{count}, which is initialized with \lstinline{1}, meaning that this recursive function is only allowed to go one level deep.

@@

    Definition HMAC_DRBG_generate_algorithm (HMAC: list Z -> list Z -> list Z) (reseed_interval: Z) (working_state: DRBG_working_state) (requested_number_of_bytes: Z) (additional_input: list Z): DRBG_generate_algorithm_result :=
        match working_state with (v, key, reseed_counter) =>
          if reseed_counter >? reseed_interval then generate_algorithm_reseed_required
          else
            let (key, v) := match additional_input with
                              | [] => (key, v)
                              | _::_ => HMAC_DRBG_update HMAC additional_input key v
                            end in
            let (v, temp) := HMAC_DRBG_generate_helper_Z HMAC key v requested_number_of_bytes in
            let returned_bits := firstn (Z.to_nat requested_number_of_bytes) temp in
            let (key, v) := HMAC_DRBG_update HMAC additional_input key v in
            let reseed_counter := reseed_counter + 1 in
            generate_algorithm_success returned_bits (v, key, reseed_counter)
        end.
\end{lstlisting}

@@

Here, \lstinline{HMAC_Final} is the name of the \lstinline{final} function verified by previous efforts, and\\ \lstinline{mbedtls_md_hmac_finish} is the name mbedTLS expects. The API specification (how the function interacts with local variable and memory) for this function is almost a complete copy of the specification defined by the previous HMAC verification efforts. The proofs that the wrapper functions satisfies the API specifications is also simple.

@@

Recall that in the functional specification (\S \ref{funcstate}), I chose to represent the DRBG state as the following:

@@

When writing the API specification, I chose a representation that more closely resembles what is in \stdtitle{mbedTLS}. Naturally, \stdtitle{mbedTLS} represents the DRBG state as a C struct. Note that the original \stdtitle{mbedTLS} code includes a function pointer for the entropy function as well as context to pass to the entropy function. As noted earlier, for the purposes of this project, I made the modifications to remove this function pointer.

@@

Lastly, I defined the conversions between the abstract state and the functional state:

@@

One of the goals of my project is to serve as a bleeding-edge adopter of the VST framework and the HMAC API specification. As is the nature of bleeding-edge adoption, I encountered many difficulties, such as a proof tactic failing silently and Verifiable C's theory of pointer comparisons being too weak. These difficulties are fixed as they get reported back to the developers of VST. In this section, I highlight one particular difficulty with the HMAC API specification.

\subsection{HMAC API Specification} \label{oldhmac}

@@

This representation works in the sense that the verification of the \stdtitle{OpenSSL} functions went through. Unfortunately, for clients of this API specification, this representation is often not good enough. The biggest issue is that this representation is too low-level; it does not keep track of the key. Upon being notified of this issue, the developers of the HMAC API specification were quickly able to provide a more abstract representation of the HMAC state, leading to the state machine representation shown earlier in \S \ref{hmac_interface}. The core of the verification code did not need to be changed, with some book-keeping added at the ends of translate between the new, more abstract representation to the more low-level implementation. While this issue was easily fixed, it would have been very hard to discover this issue without actually being clients of the API specification. This highlights the importance of having being the first client of the HMAC API specification.

@@

\begin{lstlisting}[language=C]
    int mbedtls_md_setup( mbedtls_md_context_t *ctx, const mbedtls_md_info_t *md_info, int hmac );

    void mbedtls_md_free( mbedtls_md_context_t *ctx );

    unsigned char mbedtls_md_get_size( const mbedtls_md_info_t *md_info );

@@

@@

    int mbedtls_md_hmac_reset( mbedtls_md_context_t *ctx );
\end{lstlisting}

@@

The \lstinline{md_info} field can mostly be ignored, as it is used to differentiate between the different types of message digest, and in this case, we are specializing the message digest to use HMAC and SHA-256. In order for clients of the message digest interface to have access to the \stdtitle{OpenSSL} HMAC context, I chose to have the \lstinline{hmac_ctx} pointer point to the \stdtitle{OpenSSL} HMAC context. This initialization happens in the \lstinline{mbedtls_md_setup} function:

@@

Lastly, the \lstinline{WITH} value \lstinline{kv} represents the pointer value to the SHA K vector, which is a global constant array in the SHA-256 implementation. For more information, see the verification of SHA-256 \cite{sha}.

@@

Again, much like the \lstinline{update} function, the next three \lstinline{WITH} values are the pointer value to the DRBG state, the abstract DRBG state, and the concrete DRBG state. The only new property is in the \lstinline{PROP} part of the precondition, requiring the \lstinline{entropy_len} component to be 32. Note that in \stdtitle{NIST SP 800-90A} and therefore in the functional specification, this is a global, fixed constant, but \stdtitle{mbedTLS} varies this value based on the type of HMAC used and fixes the value upon instantiation. With HMAC256, this is fixed to be 32.
